Towards a framework for reasoning about stateful compiler IRs using coinductive reasoning.

### 2024-06-25 paper.tex
shuffle projects around

Now, formal verification has been done before. Most notable is the
CompCert project: a fully verified end-to-end C compiler.~\cite{leroyCompCertFormallyVerified} Yet, the vast majority of
code today is still being compiled with non-verified compilers. We ask,
thus, how do we formally verify not \emph{a} compiler, but, how do we
scale formal verification to \emph{mainstream} compilers.

@@

We need the full power of a theorem prover to be able to capture the
full extensible nature of MLIR. Still, to recover the ease-of-use that
SMT-solver-based tools provide, we've also been investing in building
solid automation for bitvector proofs in collaboration with the Lean
community.

In the next two years, I'd like to look into a feature of MLIR called
\emph{Graph Regions}. Specific dialects may use such regions to allow a
variable to be used \emph{before} it is defined --- hence, turning the
def-use chain into a general graph. Generally, the semantics of such
dialects are statefull, and cycles are broken by looking at the value of
a variable at a previous state. For example, graph regions are used by
dialects that model hardware,
where variables model wires and cycles represent physical cycles in the
circuit.~\cite{eldridgeMLIRHardwareCompiler}

@@

The bulk of my time has been spent on developing, in collaboration with
other members of our research group, the LeanMLIR framework, culminating
in our ITP 2024 paper \emph{``Verifying Peephole Rewriting In SSA
Compiler IRs''}. As part of this development, we

@@

\begin{quote}
June '24 - January '25
\end{quote}

As mentioned in the progress report, I'm currently contributing to a
verified bitblasting procedure in Lean, called LeanSAT. Once this is
implemented, I plan to write a paper contrasting this verified
bitblaster and SAT solver with popular non-verified SMT solvers like Z3,
to investigate what exactly the cost is of the extra assurance that a
verified implementation gives. This is particularly relevant, since,
just like compilers, SMT solvers are highly complicated pieces of
software in which bugs are known to exist
\cite{brummayerFuzzingDeltadebuggingSMT2009}.

@@

LeanSAT is, however, not a silver bullet. Just like SMT solvers, it can
only reason about bitvector expressions where the width of the
bitvectors involved is bounded. One of the topics I'd like to explore
more is the connection between bitvectors and 2-adic numbers (which are
infinite and thus coinductive in nature!) to build a decision procedure
which can decide equality of a certain (limited) class of
arbitrary-width bitvector expressions.

Beyond being more satisfying, this arbitrary-width procedure is likely
more performant than SAT solving, and thus relevant even if the end-user
isn't necessarily interested in proving their property for all possible
widths.

@@

@@

\subsubsection{Runtime behaviour}\label{runtime-behaviour}
\begin{quote}
Jan '25 - May '25
\end{quote}

@@

Finally, we return to the main objective: formalizing compiler IRs.

\begin{quote}
Apr '25 - June '25
\end{quote}

Therefore, I'd like to use my library for coinductive types to define a
coinductive monad which supports computable infinite behaviours, drawing
inspiration from the interaction tree library for coq.
\cite{xiaInteractionTreesRepresenting2020} This addresses a
weakness in LeanMLIR while simultaneously giving a nice case study for
the usability of the coinduction library.

\begin{quote}
June '25 - Sep '25
\end{quote}

Here, \texttt{y} is obtained as the bitwise and of the argument
\texttt{x} and \texttt{y} itself. As a program, this doesn't make much
sense, but this style of definition is common in hardware
representations, where this is interpreted as: ``the value of \texttt{y}
in the \emph{current} clock cycle is computed by taking the bitwise and
of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''.~\footnote{
  This example is simplified, in reality you'd have to specify a
  register instruction to control when a value is stored across clock
  cycles. See https://circt.llvm.org/ for more details}
To formally model this, we could denote variables as infinite streams of bitvectors,
which gives us the value at every possible future clock cycle. Another
example of a dialect with graph regions is the FSM, or Finite State
Machine, dialect.

>>>>

Now, formally verified compilation has been done before.
Most notable is the CompCert project: a fully verified end-to-end C compiler.~\cite{leroyCompCertFormallyVerified} Yet, the vast majority of
code today is still being compiled with non-verified compilers. We ask,
thus, how do we formally verify not \emph{a} compiler, but, how do we
scale formal verification to \emph{mainstream} compilers.

@@

We need the full power of a theorem prover to be able to capture the
extensible and high-level nature of MLIR dialects.
Still, to recover the ease-of-use that SMT-solver-based tools provide, 
we've also been investing in building solid automation for bitvector proofs in collaboration with the Lean community.

In the next two years, I'd like to look into a feature of MLIR called
\emph{Graph Regions}. Specific dialects may use such regions to allow a
variable to be used \emph{before} it is defined --- hence, turning the
def-use chain into a general graph. Generally, the semantics of such
dialects are stateful, and cycles are broken by looking at the value of
a variable at a previous state.
For example, graph regions are used by
dialects that model hardware,
where variables model wires and cycles represent physical cycles in the
circuit.~\cite{eldridgeMLIRHardwareCompiler}

@@

The bulk of my time has been spent on developing, in collaboration with
other members of our research group, the LeanMLIR framework, culminating
in our ITP 2024 paper \emph{``Verifying Peephole Rewriting In SSA
Compiler IRs''}.

@@

\begin{quote}
June '24 - November '24
\end{quote}

As mentioned in the progress report, I'm currently contributing to a
verified bitblasting procedure in Lean, called LeanSAT.

@@

LeanSAT is, however, not a silver bullet. Just like SMT solvers, it can
only reason about bitvector expressions where the width of the
bitvectors involved is bounded --- and even then, only if the bound is not too large.
One of the topics I'd like to explore more is the connection between bitvectors and 2-adic numbers (which are
infinite and thus coinductive in nature!) to build a decision procedure
which can decide equality of a certain (limited) class of
arbitrary-width bitvector expressions.

Furthermore, during my earlier work on LeanMLIR, we've already encountered a 
few other strategies for solving other classes of arbitrary-widt bitvector problems.
So, even though deciding these problems in general is undecidable, we can identify
specific classes of problems that do have decision procedures.

I'm planning to write a paper about this work tentatively titled: "\emph{Taming the Bitvector Bestiary: A Characterization of Whitebox Bitvector Proof Strategies}", 
in which we explore all of these different strategies, 
classify exactly which class of expressions they solve, 
and investigate how many real-world problems are already covered by these strategies.
The aim is to submit this to PLDI, which has its call for papers around November.

@@

This internship will thus provide me with exactly the corpus of real-world bitvector verification problems that I need to evaluate the strategies we'll discuss in the 
aforementioned paper.

@@

\begin{quote}
December '24 - June '25
\end{quote}

Here, \texttt{y} is obtained as the bitwise and of the argument
\texttt{x} and \texttt{y} itself. As a program, this doesn't make much
sense, but this style of definition is common in hardware
representations, where this is interpreted as: ``the value of \texttt{y}
in the \emph{current} clock cycle is computed by taking the bitwise and
of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''.~\footnote{This example is simplified, in reality you'd have to specify a register instruction to control when a value is stored across clock cycles. See https://circt.llvm.org/ for more details} To formally model this, we could denote variables as infinite streams of bitvectors,
which gives us the value at every possible future clock cycle. Another
example of a dialect with graph regions is the FSM, or Finite State
Machine, dialect.

Once we've got this this, I'd like to actually write a paper about the coinduction library, using this work on graph regions as a case study.
The goal would be to submit this to POPL 25.

Finally, I'd like to explore further uses of coinduction in the formalization of compiler IRs.

\begin{quote}
June '25 - September  '25
\end{quote}

However, coinduction is actually a very good fit for this problem as well.
Therefore, I'd like to use my library for coinductive types to define a
coinductive monad which supports computable infinite behaviours, drawing
inspiration from the interaction tree library for coq.~\cite{xiaInteractionTreesRepresenting2020}
This addresses a
weakness in LeanMLIR while simultaneously giving a nice case study for
the usability of the coinduction library.

\subsubsection{Runtime behaviour}\label{runtime-behaviour}
\begin{quote}
September '25 - May '25
\end{quote}

Once we start using this interaction tree datastructure in LeanMLIR, though, 
we will likely run into performance problems if we try to actually execute our computable semantics.

@@
